A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-09-26 paper/project.tex
Changed all energy units, added arrows in rods figs, fixed tests to work with new ideal gas term, edits to project text Tested everything.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
 {We develop a classical density functional theory for
  water by combining the fundamental-measure theory functional--which
  describes hard sphere fluids--with attractive interactions based on the 
  Statistical Associating Fluid Theory (SAFT).  Using
  this approximation, we are able to describe various properties of
  water near liquid-vapor and liquid-solid interfaces, including water confined 
  to a hard-walled slit and hydrophobic rods immersed in water.  This
functional 
  will be the foundation for a efficient method for future work
  predicting the interactions of chemicals in aqueous solution on an
  atomic level.}

\tableofcontents
\end{abstract}

@@

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional,
which incorporates effects of kinetic energy.  The ideal gas
functional is given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{n(\xx)} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules.  The ideal gas free
energy functional has the property that the contact density at a hard
surface $n_c$ is proportional to the pressure on that surface,
according to the contact value theorem
\begin{equation}\label{contactvaluethm}
  p = n_c k_BT \:,
\end{equation}
which also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes.  These properties are retained by the total functional,
provided the remaining terms are \emph{purely nonlocal}. This requires
that
\begin{equation}
 \frac{\delta^2F_{\textit{excess}}[n]}{\delta n(\xx_1)~\delta n(\xx_1)}\neq\infty
\end{equation}
where $F_{\textit{excess}}[n]$ describes all the remaining terms in the functional.
This ensures that any higher order terms in the Taylor expansion
of $F[n+\Delta n]$ are very small.

@@

The association free energy for our four-site model has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx)\zeta(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of assotiation sites
\emph{not} hydrogen-bonded, and $\zeta(\xx)$ is a dimensionless
measure of the density inhomogeneity.  The fraction $X$ is determined
by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)\Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)\Delta(\xx)}
\end{align}
where the factor $\zeta$ from
Yu and We\cite{yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft} describes the inhomogeneity of the fluid
density, and reduces to unity in the homogeneous limit:
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}
\end{align}
The functional $\Delta$ which appears in the formula for $X$ is a
measure of hydrogen-bonding probability.  In the SAFT-VR
model\cite{gil-villegas-1997-SAFT-VR}, $\Delta$ is given by
\begin{align}
  \Delta(\xx) &= \kappa_\textit{assoc} g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\etadisp(\xx)) +
  \frac{1}{4}\beta\left(\frac{\delta A_1}{\delta \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well attractive
potential, and $a_1$ and $a_2$ are two terms in the dispersive term of
the free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere correlation
function, for which we use the functional of Yu and
Wu.\cite{yu2002fmt-dft-inhomogeneous-associating}
\begin{align}
  g_\sigma^{HS}(\xx) &= \frac{1}{1-\etadisp(\xx)}
  +\frac32\frac{\zeta(\xx)}{(1-\etadisp(\xx))^2}
  + \frac12\frac{\etadisp(\xx)^2\zeta(\xx)}{(1-\etadisp(\xx))^3}
\end{align}

@@

Figure \ref{fig:pressure-with-isotherms} shows the calculated pressures for
isotherms ranging from room temperature to past the critical temperature. The
solid line is the theoretical coexistence curve calculated from the SAFT
functional. For a given isotherm, the pressure curve should intersect this
coexistence curve twice, at the values of the vapor and liquid densities. The
functional matches well with the experimental data from NIST away from the
critical point. The slope of the pressure versus density curve is proportional
to the compressibility, which is close but slightly too large at room
temperature.

@@

Once we have the entropy, finding the internal energy is a simple
matter of subtraction:
\begin{align}
  U[n] &= F[n] + TS[n]
\end{align}
and the enthalpy is not much more work:
\begin{align}
  H[n] &= F[n] + TS[n] + pV
\end{align}

The calculated heat of vaporization at atmospheric pressure is
$\Delta H_{vap}$= 39.41 kJ/mol at the normal boiling point of water,
373 K. This compares well to the experimental
value measured by NIST ($\Delta H_{vap}$= 40.65 kJ/mol\cite{nistwater}), with 
an error of only a few percent.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/finding-vapor-pressure}\\
\includegraphics[width=\columnwidth]{figs/near-critical-point}
\end{center}
\caption{By plotting the free energy vs. density it is possible to
  determine a saturated liquid or vapor state by the common tangent
  method.  }
\label{fig:homogeneous}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-tension}
\end{center}
\caption{The theoretical versus experimental surface tension
  versus temperature. The experimental data is taken from NIST.\cite{nistwater}
}
\label{fig:surface-tension}
\end{figure}

@@

The majority of the empirical parameters used in our functional are
taken from the paper of Clark \emph{et al} on developing an optimal
SAFT model for water\cite{clark2006developing}.  This SAFT model
contains five empirical parameters: the hard-sphere radius, an energy
and length scale for the dispersion interaction, and an energy and
length scale for the association interaction.  In addition to the five
empirical parameters of Clark \emph{et al}, we add a single additional
parameter defining the length scale over which the density is averaged
when computing the dispersion free energy.  We use this final
parameter to fit the surface tension, with the result shown in
Figure~\ref{fig:surface-tension}.  Because the SAFT model of Clark
\emph{et al} overestimates the critical temperature---which is a
common feature of models which do not explicitly treat the critical
point---we cannot correctly describe the surface tension at all
temperatures, and chose to fit it for the temperature range at which
water is liquid at one atmosphere of pressure.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-1D}
\includegraphics[width=\columnwidth]{figs/energy-1D}
\end{center}
\caption{ Density profile (top) and 
energy densities (bottom) for a slab of water with molecules 
constrained to move in one dimension. The free energy, internal energy
and entropic energy densities are shown for a slit size of 4.0 nm. The dashed
line in the top figure represents the saturated liquid density.}
\label{fig:energy-1D}
\end{figure}

The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the \emph{exact}
density-functional theory \emph{would} include them), with the result
that the density profile will be smoothed out by CW effects. The liquid-vapor
interface is plotted in Figure \ref{fig:liquid-vapor-profile} and shows
some small oscillations.

@@

We choose a chemical potential corresponding to a pressure of 1 atm and set the
temperature to a typical room temperature of 298 K. We apply 
a constraining potential outside the slit, where we choose a small non-zero
density for numerical reasons. The slits studied range in size from 0.1 nm to 
8.0 nm. We minimize the energy and calculate density, energy, and $X$ on a
lattice
with a spacing of approximately 0.01 nm (0.2 bohrs) between lattice points.

The density profile for a slit of size 4.0 nm is shown in Figure
\ref{fig:energy-1D}. Near the interface, there are oscillations around the
liquid density which decrease near the center of the slit. We expect the center
of the slit to act approximately like bulk water, and the absence of
oscillations supports this. For larger slits (6.0 nm or more) the majority of
the water molecules in the slit act like the bulk. Based on our computations, we
also determine the smallest slit that can contain liquid water. This result is
approximately 0.6 nm. For smaller slits, the initial liquid water inside will
vaporize. The results for larger liquid-filled slits are only metastable,
however,as a vapor-filled slit is thermodynamically favorable. Based on our
calculations for a pressure of 1 atm, the work needed to pull apart
the two hard walls comprising the slit is one to two orders of magnitude less
than the free energy of the liquid-filled slit. For the range of slit sizes we
studied (0.1 to 8.0 nm), all the computational results of liquid-filled are
metastable. Lum, Chandler and Weeks \cite{lum1999hydrophobicity} discuss the
stability of water between large hydrophobic plates and the critical separation
necessary for stability. For water at room temperature and atmospheric pressure,
plates with a separation smallerthan about 100 nm will not contain stable liquid
water. They go on to say that confined water has a region of metastability down
to separations of about 5 nm. This is comparable to the slit separation where
our calculations show water inside the slit acting like the bulk liquid, which
we know to be a metastable solution. Forsman et al.\cite{forsman1996computer}
also study water between two hard walls at constant chemical potential. They 
also conclude that the liquid state is likely metastable, and there is a free
energy barrier to reach the stable vapor phase.

@@

The value of $X$ behaves as expected for water constrained to a slit. We
show in Figure \ref{fig:xassoc-1D} the number of hydrogen bonds per water 
molecule, equal to $4(1-X)$, for a slit of 4.0 nm.
Outside the slit, there is no hydrogen bonding and inside the number of hydrogen
bonds is approximately equal to 3.5, which is expected for saturated liquid
water.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ Energy versus radius for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension at room temperature, and it agrees well with the surface tension in
Figure \ref{fig:surface-tension}. }
\label{fig:energy-vs-diameter}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-vs-radius}
\end{center}
\caption{ Density profiles for 
a single hydrophobic rod with a radius of 0.5 nm using several different 
lattice spacing resolutions. We used a high resolution of 0.05 bohrs ($\approx$
0.003 nm), 
medium 0.2 bohrs ($\approx$ 0.001 nm), and low 0.5 bohrs ($\approx$ 0.03 nm).}
\label{fig:densityresolution}
\end{figure}

\subsection{One hydrophobic rod}
We move into two dimensions by studying a single hydrophobic rod
immersed in water. Figure \ref{fig:density-single-rod} shows density
profiles for three different radii, 0.3 nm, 0.5 nm, and 1 nm. There are the
expected oscillations outside the rod and as the distance away from the rod
increases
the density approaches that of the bulk saturated liquid density. The small
radius
rod has comparatively larger oscillations and a higher contact density.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/xassoc-single-rod}
\end{center}
\caption{ Number of bonds per molecule near the interface with a single
hydrophobic rod. Compare with Walther\cite{walther2001carbon} fig 9. }
\label{fig:xassoc-single-rod}
\end{figure}

Figure \ref{fig:energy-vs-diameter} shows the free energy per area versus
radius for the range of rods studied, with radii from 0.05 nm to 1.5 nm. After 
about 0.5 nm or so, this reaches
an asymptote equal to the surface tension. From the inset in Figure 
\ref{fig:surface-tension}, it is clear that the surface tension at room 
temperature is approximately 72 mN/m. This is very close to the value from
the larger diameter single rod energy per area, which is about 75~mN/m. 
\needsworknow{Discussion of Walther 2001 paper\cite{walther2001carbon}}

\subsubsection{Resolution and convergence tests}
Figure \ref{fig:densityresolution} shows the density profiles for three
different values of the lattice spacing resolution. The density oscillations
around a single rod of radius 0.5 nm look nearly identical for all resolutions
except for the first peak. This contact density is important as it will be
proportional to the pressure on the rod (see equation \ref{contactvaluethm}). The contact
density calculated using the medium resolution is within less than 4\% of the
value at high resolution, while the low resolution contact density is off by
about 5\%. There is some uncertainty in the value of contact density due to the
different lattice spacings. The total free energy calculations compare much more
favorably. The medium resolution is within 0.5\% and the low resolution is
within 1\% of the high resolution free energy.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-rods-in-water}
\end{center}
\caption{ Density profiles illustrating the transition from vacuum 
to liquid water between the rods. The top figure is at a separation of 0.2nm and the
bottom is 0.4nm. }
\label{fig:density-rods}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two hydrophobic rods ranging in radius from
0.3nm to 1.2nm. All were arbitrarily set to go to zero at large separations. The
discontinuous transition corresponds to the phase change from
vacuum to liquid between the rods. }
\label{fig:rods-energy-vs-distance}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-rods}
\end{center}
\caption{ Free energy, internal energy, and the entropic energy contribution versus
separation for two hydrophobic rods of radii 0.5nm. All energies are arbitrarily
set to go to zero at large separations. }
\label{fig:energy-rods}
\end{figure}

We now look at the more interesting problem of two rods in water. We continue to
use hydrophobic rods of various radii at separation $d$ (see Figure
\ref{fig:rods}), ranging from $d=0$ (touching rods) to about 1nm. At small
separations ($d<$2nm-4nm) there is only vacuum between the rods. As the rods are
pulled apart, the vacuum region expands until some critical separation is
reached and liquid water fills the region between the rods. Figure
\ref{fig:density-rods} shows density profiles before and after this transition
for rods of radius 0.5nm. Since we are looking at only hard-walled
non-interacting rods, the critical separation for this transition will be
smaller than any system where there is an attractive force between rods. Walther
\emph{et al.}\cite{walther2004hydrodynamic} study the interactions between two
carbon nanotubes. They use molecular dynamics with Lennard-Jones potentials for
nanotubes of diameter 1.25nm and separations ranging from about 0.3nm to 1.5nm.
They also find a drying transition occurs, and find it to be at separations less
than 0.9-1.0nm\cite{walther2004hydrodynamic}. This is several times larger than
what we find for hydrophobic rods, as expected due to the carbon-carbon
interations between two nanotubes. The density oscillations are very similar to
those for the single hydrophobic rod in water, as seen in Figure
\ref{fig:density-single-rod}.

The free energy per length is calculated for several different radii and is plotted
in Figure \ref{fig:rods-energy-vs-distance}. Each data set is adjusted such that the
free energy per length goes to zero at large separations. Once a certain separation is
reached, the rods will simply act like the single rod in water and the energy will
reach an asymptote proportional to the surface tension (as in Figure \ref{fig:energy-vs-diameter}).

For one set of rods (radii of 0.5nm) the free energy is plotted with the internal energy
and the entropic energy contribution, where all are in units of energy per length (see Figure
\ref{fig:energy-rods}).

Here we will want to cite~\cite{walther2004hydrodynamic}, which gives simulations for 
two interacting
nanotubes in water.  It's not precisely the same, but if we can't find something
more similar, this would be worth citing - they use CNTS with diameter 1.25 nm
and find drying occurs for separations less than 0.9-1nm.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{We develop a classical density functional theory for water by combining the fundamental-measure theory functional--which describes hard sphere fluids--with attractive interactions based on the Statistical Associating Fluid Theory (SAFT). Using this approximation, we are able to describe the density, free energy, and
entropy of water, as well as the number of hydrogen bonds formed near liquid-vapor and liquid-solid interfaces. These interfaces include water confined to a hard-walled slit, a single hydrophobic rod immersed in water, and two hydrophobic rods 
at various separations. We found \needsworknow{conclusion-like stuff about two rods, vapor-liquid transition between rods, constant force of twice surface tension, more?}.
This functional provides a foundation for future work predicting the interactions of chemicals in aqueous solution on an
atomic level.}

%\tableofcontents
\end{abstract}

@@

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional,
which incorporates effects of kinetic energy.  The ideal gas
functional is given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the quantum 
concentration. This is the particle concentration where quantum effects
become appreciable and is given by
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}
\end{equation}
where $m$ is the molecular mass of water. The ideal gas free
energy functional has the property that the contact density at a hard
surface $n_c$ is proportional to the pressure on that surface,
according to the contact value theorem
\begin{equation}\label{contactvaluethm}
  p = n_c k_BT \:,
\end{equation}
which also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes.  These properties are retained by the total functional,
provided the remaining terms are \emph{purely nonlocal}. This requires
that
\begin{equation}
 \frac{\delta^2F_{\textit{excess}}[n]}{\delta n(\xx_1)~\delta n(\xx_1)}\neq\infty
\end{equation}
where $F_{\textit{excess}}[n]$ describes all the remaining terms in the functional.
This ensures that any higher order terms in the Taylor expansion
of $F[n+\Delta n]$ are very small.

@@

The association free energy for our four-site model has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx)\zeta(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of assotiation sites
\emph{not} hydrogen-bonded, and $\zeta(\xx)$ is a dimensionless
measure of the density inhomogeneity.  The fraction $X$ is determined
by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)\Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)\Delta(\xx)}
\end{align}
where the factor $\zeta$ from
Yu and Wu\cite{yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft} describes the inhomogeneity of the fluid
density, and reduces to unity in the homogeneous limit:
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}
\end{align}
The functional $\Delta$ which appears in the formula for $X$ is a
measure of hydrogen-bonding probability.  In the SAFT-VR
model\cite{gil-villegas-1997-SAFT-VR}, $\Delta$ is given by
\begin{align}
  \Delta(\xx) &= \kappa_\textit{assoc} g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\etadisp(\xx)) +
  \frac{1}{4}\beta\left(\frac{\delta A_1}{\delta \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well attractive
potential, and $a_1$ and $a_2$ are two terms in the dispersive term of
the free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere correlation
function, for which we use the functional of Yu and
Wu.\cite{yu2002fmt-dft-inhomogeneous-associating}
\begin{align}
  g_\sigma^{HS}(\xx) &= \frac{1}{1-\etadisp(\xx)}
  +\frac32\frac{\zeta(\xx)}{(1-\etadisp(\xx))^2}
  + \frac12\frac{\etadisp(\xx)^2\zeta(\xx)}{(1-\etadisp(\xx))^3}
\end{align}

@@

Figure \ref{fig:pressure-with-isotherms} shows the calculated pressures for
isotherms ranging from room temperature to past the critical temperature. The
solid line is the theoretical coexistence curve calculated from the SAFT
functional. For a given isotherm, the pressure curve should intersect this
coexistence curve twice, at the values of the vapor and liquid densities. Even
though the theoretical pressure isotherms drop below the coexistence curve,
this is non-physical and will not actually occur. The experimental data shows
how the pressure will jump to the other side of the coexistence curve (for
temperatures less than the critical temperature), corresponding to a phase 
change in the material. The
functional matches well with the experimental data from NIST away from the
critical point. The slope of the pressure versus density curve is proportional
to the compressibility, which is close but slightly too large at room
temperature.

@@

Once we have the entropy, finding the internal energy is simple:
\begin{align}
  U[n] &= F[n] + TS[n]
\end{align}
and the enthalpy is not much more work:
\begin{align}
  H[n] &= F[n] + TS[n] + pV
\end{align}

The calculated heat of vaporization at atmospheric pressure is
$\Delta H_{vap}$= 39.41~kJ/mol at the normal boiling point of water,
373 K. This compares well to the experimental
value measured by NIST ($\Delta H_{vap}$= 40.65~kJ/mol\cite{nistwater}), with 
an error of only a few percent.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/finding-vapor-pressure}\\
\includegraphics[width=\columnwidth]{figs/near-critical-point}
\end{center}
\caption{By plotting the free energy versus density it is possible to
  determine a saturated liquid or vapor state by the common tangent
  method. The top plot illustrates the common tangents near room 
  temperature, while the bottom plot shows the same thing close to the 
  critical temperature. }
\label{fig:homogeneous}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-tension}
\end{center}
\caption{The theoretical versus experimental surface tension
  versus temperature. The experimental data is taken from NIST.\cite{nistwater}
  The length-scaling parameter $\sigma$ is fit so that the theoretical surface tension
  will match the experimental surface tension near room temperature.
}
\label{fig:surface-tension}
\end{figure}

@@

The majority of the empirical parameters used in our functional are
taken from the paper of Clark \emph{et al} on developing an optimal
SAFT model for water\cite{clark2006developing}.  This SAFT model
contains five empirical parameters: the hard-sphere radius, an energy
and length scale for the dispersion interaction, and an energy and
length scale for the association interaction.  In addition to the five
empirical parameters of Clark \emph{et al}, we add a single additional
parameter $\sigma$ defining the length scale over which the density is averaged
when computing the dispersion free energy.  We use this final
parameter to fit the surface tension, with the result shown in
Figure~\ref{fig:surface-tension}.  Because the SAFT model of Clark
\emph{et al} overestimates the critical temperature---which is a
common feature of models which do not explicitly treat the critical
point---we cannot correctly describe the surface tension at all
temperatures, and chose to fit it for the temperature range at which
water is liquid at one atmosphere of pressure.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-1D}
\includegraphics[width=\columnwidth]{figs/energy-1D}
\end{center}
\caption{ Density profile (top) and 
energy densities (bottom) for a slab of water with molecules 
constrained to move in one dimension. The free energy, internal energy
and entropic energy densities are shown for a slit size of 4.0~nm. The dashed
line in the top figure represents the saturated liquid density.}
\label{fig:energy-1D}
\end{figure}

The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the \emph{exact}
density-functional theory \emph{would} include them), with the result
that the density profile will be smoothed out by CW effects. The liquid-vapor
interface is plotted in Figure~\ref{fig:liquid-vapor-profile} and shows
some small oscillations.

@@

We choose a chemical potential corresponding to a pressure of 1~atm and set the
temperature to a typical room temperature of 298~K. We apply 
a constraining potential outside the slit, where we choose a small non-zero
density for numerical reasons. The slits studied range in size from 0.1~nm to 
8.0~nm. We minimize the energy and calculate density, energy, and $X$ on a
lattice
with a spacing of approximately 0.01~nm (0.2~bohrs) between lattice points.

The density profile for a slit of size 4.0~nm is shown in Figure~\ref{fig:energy-1D}. Near the interface, there are oscillations around the
liquid density which decrease near the center of the slit. We expect the center
of the slit to act approximately like bulk water, and the absence of
oscillations supports this. For larger slits (6.0~nm or more) the majority of
the water molecules in the slit act like the bulk. Based on our computations, we
also determine the smallest slit that can contain liquid water. This result is
approximately 0.6~nm. For smaller slits, the initial liquid water inside will
vaporize. The results for larger liquid-filled slits are only metastable,
however,as a vapor-filled slit is thermodynamically favorable. Based on our
calculations for a pressure of 1~atm, the work needed to pull apart
the two hard walls comprising the slit is one to two orders of magnitude less
than the free energy of the liquid-filled slit. For the range of slit sizes we
studied (0.1 to 8.0~nm), all the computational results of liquid-filled are
metastable. Lum, Chandler and Weeks \cite{lum1999hydrophobicity} discuss the
stability of water between large hydrophobic plates and the critical separation
necessary for stability. For water at room temperature and atmospheric pressure,
plates with a separation smaller than about 100~nm will not contain stable liquid
water. They go on to say that confined water has a region of metastability down
to separations of about 5~nm. This is comparable to the slit separation where
our calculations show water inside the slit acting like the bulk liquid, which
we know to be a metastable solution. Forsman \emph{et al.}\cite{forsman1996computer}
also study water between two hard walls at constant chemical potential. They 
also conclude that the liquid state is likely metastable, and there is a free
energy barrier to reach the stable vapor phase.

@@

The value of $X$ behaves as expected for water constrained to a slit. We
show in Figure~\ref{fig:xassoc-1D} the number of hydrogen bonds per water 
molecule, equal to $4(1-X)$, for a slit of 4.0~nm.
Outside the slit, there is no hydrogen bonding and inside the number of hydrogen
bonds is approximately equal to 3.5, which is expected for saturated liquid
water.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ Energy versus radius for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension at room temperature, and it agrees well with the surface tension in
Figure~\ref{fig:surface-tension}. }
\label{fig:energy-vs-diameter}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/xassoc-single-rod}
\end{center}
\caption{ Number of bonds per molecule near the interface with a single
hydrophobic rod. There is a decrease in the hydrogen bonding near the surface 
of the rod, as expected at the interface of any hydrophobic wall. }
\label{fig:xassoc-single-rod}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-vs-radius}
\end{center}
\caption{ Density profiles for 
a single hydrophobic rod with a radius of 0.5~nm using several different 
lattice spacing resolutions. We used a high resolution of 0.05~bohrs ($\approx$
0.003~nm), 
medium 0.2~bohrs ($\approx$ 0.001~nm), and low 0.5~bohrs ($\approx$ 0.03~nm).}
\label{fig:densityresolution}
\end{figure}

\subsection{One hydrophobic rod}
We move into two dimensions by studying a single hydrophobic rod
immersed in water. Figure~\ref{fig:density-single-rod} shows density
profiles for three different radii, 0.3~nm, 0.5~nm, and 1~nm. There are the
expected oscillations outside the rod and as the distance away from the rod
increases
the density approaches that of the bulk saturated liquid density. The small
radius
rod has comparatively larger oscillations and a higher contact density.

Figure~\ref{fig:energy-vs-diameter} shows the free energy per area versus
radius for the range of rods studied, with radii from 0.05~nm to 1.5~nm. After 
about 0.5~nm or so, this reaches
an asymptote equal to the surface tension. From the inset in Figure~\ref{fig:surface-tension}, the surface tension at room 
temperature is approximately 72~mN/m. This is very close to the value from
the larger diameter single rod energy per area, which is about 75~mN/m.

The value of $X$ is calculated and converted to number of bonds per molecule and 
shown for a rod with radius 0.2~nm in Figure~\ref{fig:xassoc-single-rod}. Inside the
rod (less than 0.2~nm) there are no molecules to form hydrogen bonds although the value
of $X$ is still calculated there. Walther \emph{et al.}\cite{walther2001carbon} 
calculate the number of hydrogen bonds for a carbon nanotube immersed in water.
They show a decrease in the bonding from a bulk value of 3.75 to 2.89 near the 
interface with the nanotube. This is a much larger change than for our hydrophobic
rod of a similar radius, where the number of bonds decreases from a bulk value of 3.5 
to about 3.25 at the interface (see Figure~\ref{fig:xassoc-single-rod}).

\subsubsection{Resolution and convergence tests}
Figure~\ref{fig:densityresolution} shows the density profiles for three
different values of the lattice spacing resolution. The density oscillations
around a single rod of radius 0.5~nm look nearly identical for all resolutions
except for the first peak. This contact density is important as it will be
proportional to the pressure on the rod (see Equation~\ref{contactvaluethm}). The contact
density calculated using the medium resolution is within less than 4\% of the
value at high resolution, while the low resolution contact density is off by
about 5\%. There is some uncertainty in the value of contact density due to the
different lattice spacings. The total free energy calculations compare much more
favorably. The medium resolution is within 0.5\% and the low resolution is
within 1\% of the high resolution free energy.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-rods-in-water}
\end{center}
\caption{ Density profiles illustrating the transition from vapor 
to liquid water between the rods. The radius is 0.5~nm, the top figure is at a separation of 0.2~nm and the
bottom is 0.4~nm. }
\label{fig:density-rods}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two hydrophobic rods ranging in radius from
0.3~nm to 0.9~nm. All were arbitrarily set to go to zero at large separations. The
discontinuous transition corresponds to the phase change from
vapor to liquid between the rods. }
\label{fig:rods-energy-vs-distance}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-rods}
\end{center}
\caption{ Free energy, internal energy, and the entropic energy contribution versus
separation for two hydrophobic rods of radii 0.5~nm. All energies are arbitrarily
set to go to zero at large separations. }
\label{fig:energy-rods}
\end{figure}

We now look at the more interesting problem of two rods in water. We continue to
use hydrophobic rods of various radii at separation $d$ (see Figure~\ref{fig:rods}), ranging from $d=0$ 
(touching rods) to about 1~nm. At small
separations ($d<2$~-~4~nm) there is only vapor between the rods. As the rods are
pulled apart, the vapor region expands until some critical separation is
reached and liquid water fills the region between the rods. Figure~\ref{fig:density-rods} 
shows density profiles before and after this transition
for rods of radius 0.5~nm. Since we are looking at only hard-walled
non-interacting rods, the critical separation for this transition will be
different than any system where there is an appreciable force between rods.

The free energy per length is calculated for several different radii and is plotted
in Figure~\ref{fig:rods-energy-vs-distance}. Each data set is adjusted such that the
free energy per length goes to zero at large separations. Once a certain separation is
reached, the rods will simply act like the single rod in water and the energy will
reach an asymptote proportional to the surface tension (as in Figure~\ref{fig:energy-vs-diameter}).

For one set of rods (radii of 0.5~nm) the free energy is plotted with the internal energy
and the entropic energy contribution, where all are in units of energy per length 
(see Figure~\ref{fig:energy-rods}). \needsworknow{More discussion of Figure~\ref{fig:energy-rods}}

Walther
\emph{et al.}\cite{walther2004hydrodynamic} study the interactions between two
carbon nanotubes. They use molecular dynamics with Lennard-Jones potentials for
nanotubes of diameter 1.25~nm and separations ranging from about 0.3~nm to 1.5~nm.
They also find a drying transition occurs, and find it to be at separations less
than 0.9~-~1.0~nm\cite{walther2004hydrodynamic}. This is several times larger than
what we find for hydrophobic rods, as expected due to the carbon-carbon
interations between two nanotubes.

The density oscillations are very similar to
those for the single hydrophobic rod in water, as seen in Figure~\ref{fig:density-single-rod}.
